Large and Infinitary Quotient Inductive-Inductive Types\thanks{This work was supported by EFOP-3.6.3-VEKOP-16-2017-00002 grant and COST Action EUTypes CA15123.}

### 2019-12-26 draft/old.tex
work on paper

\input{abbrevs.tex}
\newcommand{\Lift}{\Uparrow}
\newcommand{\ToS}{\mathsf{ToS}}
\newcommand{\ext}{\triangleright}
\newcommand{\emptycon}{\scaleobj{.75}\bullet}
\newcommand{\Pii}{\mathsf{\Pi}}
\newcommand{\appi}{\mathsf{app}}
\newcommand{\lami}{\mathsf{lam}}
\newcommand{\Pie}{\mathsf{\hat{\Pi}}}
\newcommand{\appe}{\mathsf{ap\hat{p}}}
\newcommand{\lame}{\mathsf{la\hat{m}}}
\newcommand{\Piinf}{\mathsf{\tilde{\Pi}}}
\newcommand{\appinf}{\mathsf{ap\tilde{p}}}
\newcommand{\laminf}{\mathsf{la\tilde{m}}}

\newcommand{\appitt}{\mathop{{\scriptstyle @}}}

\newcommand{\Refl}{\mathsf{Refl}}

\newcommand{\IdU}{\mathsf{IdU}}
\newcommand{\ReflU}{\mathsf{ReflU}}
\newcommand{\Sig}{\mathsf{Sig}}

\newcommand{\Sg}{\mathsf{\Sigma}}
\newcommand{\flCwF}{\mathsf{flCwF}}
\newcommand{\Kfam}{\mathsf{K}}

@@

\title{Generalizations of Quotient Inductive-Inductive Types\thanks{This work
    was supported by EFOP-3.6.3-VEKOP-16-2017-00002 grant and COST
    Action EUTypes CA15123.}}

@@

\begin{abstract}
Quotient inductive-inductive types (QIITs) are generalized inductive types which
allow sorts to be indexed over previously declared sorts, and allow usage of
equality constructors. QIITs are especially useful for algebraic descriptions of
type theories and constructive definitions of real, ordinal and surreal
numbers. We develop new metatheory for large QIITs, large elimination, recursive
equations, infinitary constructors and equations between sorts.
%% First, we develop tools which allow precise and convenient handling
%% of universe levels, working internally in an extensional type theory with
%% cumulative universes.
As in prior work, we describe QIITs using a type theory where each context
represents a QIIT signature. However, in our case the theory of signatures can
also describe its own signature. We use self-description to bootstrap a model
theory for the theory of signatures without using preterms or assuming a
pre-existing internal syntax for a type theory. We give initial algebra
semantics for described QIITs, and we show the equivalence of initiality and
induction. We present two extensions of a previous term model construction. The
first one constructs all large infinitary QIITs without sort equations from the
QIIT of the theory of signatures. The second one constructs all large finitary
QIITs with sort equations, from the same syntax. This separation is required
because handling infinitary constructors in the term model requires showing a
strong form of invariance under algebra isomorphisms, which is violated by sort
equations.

\end{abstract}

@@

ISSUE. We actually need flCwF pseudomorphisms for Sub in infinitary semantics.
$a : \Tm\,\Gamma\,\U$ is \emph{not} a discrete displayed flCwF, only $\El\,a$ is!
$\Piinf\,A\,b : \Tm\,\Gamma,\U$ does not preserve (displayed) object in $\Gamma$
strictly.

\begin{itemize}
\item New semantics with pseudomorphisms for Sub and Tm. Is it enough to
      have weak preservation of comprehension, and we get all other preservation
      for free, see ``gluing for type theory''?
\item Check term model and eliminators to see if change needed.
\item Check if the isomorphism for K in flCwF can be made strict, i.e.
      $\Sub\,\Gamma\,\Delta = \Tm\,\Gamma\,(\Kfam\,\Delta)$.
\end{itemize}

@@

The main motivation of the current paper is to extend the notion of QIIT so that
it accommodates all algebraic notions of type theories which have been used in
previous works. As a side effect of fulfilling this goal, infinitary QIITs such
as Cauchy real numbers are covered as well.

We generalize previous notions of QIITs in the following ways:
\begin{enumerate}
  \item
  \textbf{Large constructors, large elimination} and models at different
  universe levels. This feature is routinely used in the metatheory of type
  theory, but it has not been presented explicitly in previous works about
  QIITs.
  \item
  \textbf{Infinitary constructors}. This allows infinitely branching
  trees. Also, the theory of QIIT signatures is itself large and infinitary,
  thus now it can ``eat itself'', i.e.\ include its own signature and provide
  its own metatheory. This was previously not possible in \cite{TODO}, where
  only finitary QIITs were described. We use this self-representation to
  bootstrap the model theory of signatures, without having to assume any
  pre-existing internal syntax.
  \item
  \textbf{Recursive equations}, i.e. equations appearing as assumptions
  of constructors. These have occurred previously in syntaxes of cubical type
  theories, as boundary conditions\cite{TODO}.
  \item
  \textbf{Sort equations}, or equations between type constructors. They have
  been used recently to conveniently represent type-theoretic universes, for
  example Russell-style universes\cite{TODO} or cumulative hierarchies\cite{TODO}.
\end{enumerate}

>>>>

\input{abbrevs.tex}
\newcommand{\Lift}{\Uparrow}
\newcommand{\ToS}{\mathsf{ToS}}
\newcommand{\ext}{\triangleright}
\newcommand{\emptycon}{\scaleobj{.75}\bullet}
\newcommand{\Pii}{\mathsf{\Pi}}
\newcommand{\appi}{\mathsf{app}}
\newcommand{\lami}{\mathsf{lam}}
\newcommand{\Pie}{\mathsf{\hat{\Pi}}}
\newcommand{\appe}{\mathsf{ap\hat{p}}}
\newcommand{\lame}{\mathsf{la\hat{m}}}
\newcommand{\Piinf}{\mathsf{\tilde{\Pi}}}
\newcommand{\appinf}{\mathsf{ap\tilde{p}}}
\newcommand{\laminf}{\mathsf{la\tilde{m}}}
\newcommand{\appitt}{\mathop{{\scriptstyle @}}}
\newcommand{\Refl}{\mathsf{Refl}}
\newcommand{\IdU}{\mathsf{IdU}}
\newcommand{\ReflU}{\mathsf{ReflU}}
\newcommand{\Sig}{\mathsf{Sig}}
\newcommand{\Sg}{\mathsf{\Sigma}}
\newcommand{\flCwF}{\mathsf{flCwF}}
\newcommand{\Kfam}{\mathsf{K}}

@@

\title{Large and Infinitary Quotient Inductive-Inductive Types\thanks{This work
    was supported by EFOP-3.6.3-VEKOP-16-2017-00002 grant and COST
    Action EUTypes CA15123.}}

@@

\begin{abstract}
Quotient inductive-inductive types (QIITs) are generalized inductive types which
allow sorts to be indexed over previously declared sorts, and allow usage of
equality constructors. QIITs are especially useful for algebraic descriptions of
type theories and constructive definitions of real, ordinal and surreal
numbers. We develop new metatheory for large QIITs, large elimination, recursive
equations and infinitary constructors. As in prior work, we describe QIITs using
a type theory where each context represents a QIIT signature. However, in our
case the theory of signatures can also describe its own signature, modulo
universe sizes. We bootstrap the model theory of signatures using
self-description and a Church-coded notion of signature, without using
complicated raw syntax or assuming an existing internal QIIT of signatures. We
give semantics to described QIITs by modeling each signature as a finitely
complete CwF (category with families) of algebras. Compared to the case of
finitary QIITs, we additionally need to show invariance under algebra
isomorphisms in the semantics. We do this by modeling signature types as
isofibrations. We show by a term model construction that every QIIT is
constructible from the syntax of the theory of signatures.
\end{abstract}

@@

%% ISSUE. We actually need flCwF pseudomorphisms for Sub in infinitary semantics.
%% $a : \Tm\,\Gamma\,\U$ is \emph{not} a discrete displayed flCwF, only $\El\,a$ is!
%% $\Piinf\,A\,b : \Tm\,\Gamma,\U$ does not preserve (displayed) object in $\Gamma$
%% strictly.

%% \begin{itemize}
%% \item New semantics with pseudomorphisms for Sub and Tm. Is it enough to
%%       have weak preservation of comprehension, and we get all other preservation
%%       for free, see ``gluing for type theory''?
%% \item Check term model and eliminators to see if change needed.
%% \item Check if the isomorphism for K in flCwF can be made strict, i.e.
%%       $\Sub\,\Gamma\,\Delta = \Tm\,\Gamma\,(\Kfam\,\Delta)$.
%% \end{itemize}

@@

The main motivation of the current paper is to extend the notion of QIIT so that
it accommodates more algebraic notions of type theories than previously, and
more aspects of their practical metatheory. As a side effect of fulfilling this
goal, infinitary QIITs such as Cauchy real numbers are covered as well.

We generalize previous notions of QIITs in the following ways:
\begin{enumerate}
  \item
  \textbf{Large constructors, large elimination} and models at different
  universe levels. Large elimination is routinely used in the metatheory of type
  theory, but it has not been presented explicitly in previous works about
  QIITs.
  \item
  \textbf{Infinitary constructors}. This allows infinitely branching
  trees. Also, the theory of QIIT signatures is itself large and infinitary,
  thus now it can ``eat itself'', i.e.\ include its own signature and provide
  its own metatheory. This was not possible previously in
  \cite{kaposi2019constructing}, where only finitary QIITs were described. We
  exploit self-representation to bootstrap the model theory of signatures,
  without having to assume any pre-existing internal syntax.
  \item
  \textbf{Recursive equations}, i.e. equations appearing as assumptions
  of constructors. These have occurred previously in syntaxes of cubical type
  theories, as boundary conditions\cite{TODO}.
\end{enumerate}
